# TWO INDEPENDENT SAMPLES t-TEST: for the Difference in MEANS


Using the `t.test()` function



```{r, message=FALSE, error=FALSE}
library(car)          # Companion for Applied Regression (and ANOVA)
```




## Exploratory Data Analysis: i.e. the eyeball method

**Do the two groups, treatment and control, have different oral conditions at initial observation?  What about four weeks later?**

Judge any difference in centers (means) within the context of the within group spread (stadard deviation/variance)



### Means and SDs

```{r}
cancer_clean %>% 
  dplyr::group_by(trt) %>% 
  furniture::table1(totalcin, totalcw4,
                    na.rm = FALSE)
```


### Stacked Histograms

#### Baseline Oral Condition

```{r}
cancer_clean %>% 
  ggplot(aes(totalcin)) +
  geom_histogram(bins = 10) +
  facet_grid(trt ~ .)
```


#### Four Weeks Oral Condition

```{r}
cancer_clean %>% 
  ggplot(aes(totalcw4)) +
  geom_histogram(bins = 10) +
  facet_grid(trt ~ .)
```


### Side-by-Side Boxplots

#### Baseline Oral Condition

```{r}
cancer_clean %>% 
  ggplot(aes(x = trt,
             y = totalcin)) +
  geom_boxplot()
```



#### Four Weeks Oral Condition

```{r}
cancer_clean %>% 
  ggplot(aes(x = trt,
             y = totalcw4)) +
  geom_boxplot()
```

## Assumptions

### Independence

BOTH Samples were drawn **INDEPENDENTLY** at random (at least as representative as possible)

* Nothing can be done to fix NON-representative samples!     
* Can not for with any statistically test
* If idenpendence is violated, you may want to use a paired-samples t-test 



### Normality

> A variable is said to follow the normal distribution if it resembles the normal curve.  Specifically it is symetrical, unimodal, and bell shaped.

The continuous variable has a **NORMAL** distribution in BOTH populations

* Not as important if the sample is large *(Central Limit Theorem)*
* IF the sample is far from normal &/or small, might want to use a different method  

**Options to judging normality:**

1. Visualization of each sample's distribution    
    + Stacked **histograms**, but is sensitive to binning choices *(number or width)*
    + Side-by-side **boxplots**, shows *median* instead of *mean* as central line
    + Seperate **QQ plots** (straight $45^\circ$ line), but is sensitive to outliers!    
    
2. Calculate **Skewness** and **Kurtosis**, within each group 
    + Divided each value by its standard error (SE)    
    + A result $\gt \pm 2$ indicates issues  

3. Formal Inferencial Tests for Normality, on each group   
    + Null-hypothesis: population is normally distributed
    + A $p \lt .05$ ???indicate snon-normality
    + For smaller samples, use **Shapiro-Wilk's Test** 
    + For larger samples, use **Kolmogorov-Smirnov's Test**


  

#### Baseline Oral Condition

```{r}
cancer_clean %>% 
  ggplot(aes(sample = totalcin)) +   # make sure to include "sample = "
  geom_qq() +                        # layer on the dots
  stat_qq_line() +                   # layer on the line
  facet_grid(. ~ trt)                # panel by group
```



```{r}
cancer_clean %>% 
  dplyr::filter(trt == "Placebo") %>%   # select one group
  dplyr::pull(totalcin) %>%             # extract the continuous variable
  shapiro.test()                        # test for normality
```

```{r}
cancer_clean %>% 
  dplyr::filter(trt == "Aloe Juice") %>%    # select one group
  dplyr::pull(totalcin) %>%                 # extract the continuous variable
  shapiro.test()                            # test for normality
```


> Shapiro-Wilk's tests yield evidence that baseline oral condition is NOT normally distributed in the placebo group, W = .681, p <.001, nor the treatment group, W = .785, p = .006.  Visual inspection suggests that violatioins may by more extreme in the placebo group.


#### Four Weeks Oral Condition


```{r}
cancer_clean %>% 
  ggplot(aes(sample = totalcw4)) +   # make sure to include "sample = "
  geom_qq() +                        # layer on the dots
  stat_qq_line() +                   # layer on the line
  facet_grid(. ~ trt)                # panel by group
```



```{r}
cancer_clean %>% 
  dplyr::filter(trt == "Placebo") %>%   # select one group
  dplyr::pull(totalcw4) %>%             # extract the continuous variable
  shapiro.test()                        # test for normality
```

```{r}
cancer_clean %>% 
  dplyr::filter(trt == "Aloe Juice") %>%    # select one group
  dplyr::pull(totalcw4) %>%                 # extract the continuous variable
  shapiro.test()                            # test for normality
```


> Shapiro-Wilk's tests yielded no evidence that oral condition is NOT normally distributed four weeks after baseline in the placebo group, $W = .883, p = .064$, and the treatment group, $W = .929, p = .401$.  


### HOV

> Two Populations exhibit Homogeneity of Variance (HOV), i.e. have about the same amount of spread

Before performing the $t$ test, check to see if the assumption of homogeneity of variance is met using **Levene's Test**.  For a independent samples `t`-test for means, the groups need to have the same amount of spread (SD) in the measure of interest.

```{block type='rmdlightbulb', echo=TRUE}
Use the `car:leveneTest()` function tests the HOV assumtion.  

Inside the funtion you need to specify at least three options (sepearated by commas):

* the formula:  `continuous_var ~ grouping_var` (replace with your variable names)
* the dataset:  `data = .` to pipe it from above
* the center:   `center = "mean"` since we are comparing means
```


#### Baseline Oral Condition

Do the participants in the treatment and control groups have the same spread in oral condition at BASELINE?

```{r}
cancer_clean %>% 
  car::leveneTest(totalcin ~ trt,    # formula: continuous_var ~ grouping_var
                  data = .,          # pipe in the dataset
                  center = "mean")   # The default is "median"
```

> No violations of homogeneity were detected, $F(1, 23) = 2.210, p = .151$.



#### Four Weeks Oral Condition

Do the participants in the treatment and control groups have the same spread in oral condition at the FOURTH WEEK?

```{r}
cancer_clean %>% 
  car::leveneTest(totalcw4 ~ trt,    # formula: continuous_var ~ grouping_var
                  data = .,          # pipe in the dataset
                  center = "mean")   # The default is "median"
```

> No violations of homogeneity were detected, $F(1, 23) = 0, p = .995$.


## Inference

**Formal Statistical Test: t-Test for Difference in Independent Group Means**

```{block type='rmdlightbulb', echo=TRUE}
Use the same `t.test()` funtion we have used for a single sample, but speficy a few more options.

**Inside the funtion you need to specify at least three options (sepearated by commas):**
 
* the formula:  `continuous_var ~ grouping_var` (replace with your variable names)    
* the dataset:  `data = .` to pipe it from above    


**You MAY need/want to specify some or all of the following options you may way to leave as the default or override:**

* HOV assumed:  
    + `var.equal = FALSE` **Default** Seperate-Variance test using Welch's df    
    + `var.equal = TRUE`  Pooled-Variance test *(if HOV is NOT violated)*


* Number of tails:    
    + `alternative = "two.sided"` **Default** Allows for a 2-sided alternative 
    + `alternative = "less"` Only Allows: group 1 < group 2
    + `alternative = "greater"` Only Allows: group 1 > group 2
    
    
* Independent vs. paired:    
    + `paired = FALSE`  **Default** Conducts an INDEOENDENT groups t-Test    
    + `paired = TRUE`   Conducts a PAIRED meausres t-Test 
    
    
* Confidence level:   
    + `conf.level = 0.95` **Default** Computes the 95% confidence inverval    
    + `conf.level = 0.90`  Changes to a 90% confidence interval
```

### Pooled Variance Test 

Use when there are no violations of HOV

#### Baseline Oral Condition

Do the participants in the treatment group have a different average oral condition at BASELINE, compared to the control group?

```{r}
# Minimal syntax
cancer_clean %>% 
  t.test(totalcin ~ trt,   # formula: continuous_var ~ grouping_var
         data = .,         # pipe in the dataset
         var.equal = TRUE) # HOV was violated (option = TRUE)
```

> No evidence of a differnece in mean oral condition at baseline, $t(23) = 0.186, p = .854$.  Note: this test may be unreliable due to the non-normality of the samll samples.





#### Four Weeks Oral Condition


Do the participants in the treatment group have a different average oral condition at the FOURTH WEEK, compared to the control group?

```{r}
# Fully specified function
cancer_clean %>% 
  t.test(totalcw4 ~ trt,             # formula: continuous_var ~ grouping_var
         data = .,                   # pipe in the dataset
         var.equal = TRUE,           # default: HOV was violated (option = TRUE)
         alternative = "two.sided",  # default: 2 sided (options = "less", "greater")
         paired = FALSE,             # default: independent (option = TRUE)
         conf.level = .95)           # default: 95% (option = .9, .90, ect.)
```

> No evidence of a differnece in mean oral condition at the fourth week, $t(23) = -0.350, p = .733$.



### Seperate Variance Test 

Use if there are violations of HOV or the samples are difference sizes

