# One-Way ANOVA: Code


Required Packages 

```{r}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(furniture)    # Nice tables (by our own Tyson Barrett)
library(afex)         # Analysis of Factorial Experiments
library(emmeans)      # Estimated marginal means (Least-squares means)
library(readxl)       # Necessary for reading in an example data set
```



## Prepare for Modeling

### Ensure the Data is in "long" Format 

First, the data must be restructured from **wide** to **long** format, so that each observation is on its own line.  All categorical variables must be declared as factors.  We also must add an distinct indicator variable.  

Name of the dataset: `data_long` 

Variables:  
 
* `id_var` Name of the variable that is qunique for each unit of study (i.e. person)
* `group_IV` Name of the categorical variable (`factor`) with at least 2 levels
* `continuous_DV` Name of the continuous varaible (`dbl`) that we want to compare the averages of



### Compute Summary Statistics

Second, check the summary statistics for each of the $k$ groups.

```{r, eval=FALSE}
# Raw data: summary table
data_long %>% 
  dplyr::group_by(group_IV) %>%          # divide into groups
  furniture::table1(continuous_DV        # gives M(SD)
                    digits = 2,     
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown") 
```





### Plot the Raw Data

Third, plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.

```{r, eval=FALSE}
# Raw data: boxplots
data_long %>% 
  ggplot(aes(x = group_IV,
             y = continuous_DV)) + 
  geom_boxplot() +
  geom_point() 
```


```{r, eval=FALSE}
# Raw data: Mean-SD plots
data_long %>% 
  ggplot(aes(x = group_IV,
             y = continuous_DV)) + 
  stat_summary() 
```





### Assumption Check: HOV

It is important that we test for violations of the assumption of *Homogeneity of variance*.

```{r EX_hov, eval=FALSE}
# Levene's Test of HOV
data_name %>% 
  car::leveneTest(continuous_var ~ group_var, 
                  data = .,            
                  center = "mean")     
```



## Fitting One-way ANOVA Model


The `aov_4()` function from the `afex` package fits ANOVA models *(oneway, two-way, repeated measures, and mixed design)*. It needs at least two arguments:

1. formula:  `continuous_DV ~ group_IV + (1|id_var)`  *one observation per subject and `id_var` is distinct for each subject*

2. dataset: `data = .` *we use the period to signify that the datset is being piped from above*


Here is an outline of what your syntax should look like when you **fit and save a one-way ANOVA**.  Of course you will replace the dataset name and the variable names, as well as the name you are saving it as.

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** The `aov_4()` function works on data in LONG format only.  Each observation needs to be on its one line or row with seperate variables for the group membership (categorical factor or `fct`) and the continuous measurement (numberic or `dbl`).
```

```{r, eval=FALSE}
# One-way ANOVA: fit and save
aov_name <- data_long %>% 
  afex::aov_4(continuous_DV ~ group_IV + (1|id_var),
              data = .)
```





### Basic Output - stored name of model

By running the name you saved you model under, you will get a brief set of output, including a measure of **Effect Size**.

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** The `ges` is the *generalized eta squared*.  In a one-way ANOVA, the eta-squared effect size is only one value, ie. generalized $\eta_g$ and partial $\eta_p$ are the same.
```



```{r, eval=FALSE}
# Display basic ANOVA results (includes effect size)
aov_name 
```


### Fuller Output - add $Anova on model name

To fully fill out a standard ANOVA table and compute other effect sizes, you will need a more complete set of output, including the **Sum of Squares** components, you will need to add `$Anova` at the end of the model name before running it.

```{block type='rmdlightbulb', echo=TRUE}
 **NOTE:** IGNORE the first line that starts with `(Intercept)`!  Also, the 'mean sum of squares' are not included in this table, nor is the **Total** line at the bottom of the standard ANOVA table.  You will need to manually compute these values and add them on the homework page.  Remember that `Sum of Squares (SS)` and `degrees of freedom (df)` add up, but `Mean Sum of Squreas (MS)` do not add up.  Also: `MS = SS/df` for each term.
```

```{r, eval=FALSE}
# Display fuller ANOVA results (includes sum of squares)
aov_name$Anova 
```


## Followup-tests


### Estimated Marginal Means with `emmeans::emmeans(~ group_var)`


```{r EX_emm, eval=FALSE}
aov_name %>% 
  emmeans::emmeans(~ group_var)        # Calculate Estimated Marinal Means
```





### All Pairwise Comparisons with `pairs()`

There are two steps to conduct all possible pairwise comparisons:

1. `emmeans::emmeans(~ group_var)` - Calculate the *Estimated Marinal Means*
2. `pairs()` - Determine if each pair is *significantly different*

Within the `pairs()` function there are several options for controling for multiple comparisons, including:

* `adjust = "none"` - Fisher's LSD
* `adjust = "tukey"` - Tukey's HSD
* `adjust = "bon"` - Bonferroni


```{r EX_tukey, eval=FALSE}
# Pairwise post hoc: Tukey's HSD adjustment for multiple comparisons
aov_name %>% 
  emmeans::emmeans(~ group_var) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "tukey")                  # Is each pair signif different?
```



### Plot Marginal Means with `emmeans::emmip(~ group_var)`


```{r EX_emmip, eval=FALSE}
aov_name %>% 
  emmeans::emmip(~ group_var,        # Calculate Estimated Marginal Means
                 CIs = TRUE)         # Include Confidence Intervals
```





### Contrast Statements with `emmeans::contrast()`

There are two steps to conduct a contrast comparison:

1. `emmeans(~ group_var)` - Calculate the *Estimated Marinal Means*
2. `contrast()` - Determine if each pair is *significantly different*

Inside the contrast statement, list the named sets of linear contrast weights.  We will only be doing one-at-a-time, but we must still use a nested `list`.  
```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** You must provide one weight ($c_i$) for each of the $k$ groups.  If you wish to ignore a group, that group's weight is $c_i = 0$.  The sum total of all the weights must be zero ($\sum c_i = 0$), so use positive and negative numbers.
```

```{r EX_contrast, eval=FALSE}
# Contrast statement : Impossible vs. Rest
aov_name %>% 
  emmeans::emmeans(~ group_var) %>% 
  emmeans::contrast(list("your contrast name" = c(c_1, c_2, ... , c_k)))
```