# One-Way ANOVA: Code


Required Packages 

```{r}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(furniture)    # Nice tables (by our own Tyson Barrett)
library(afex)         # Analysis of Factorial Experiments
library(emmeans)      # Estimated marginal means (Least-squares means)
library(readxl)       # Necessary for reading in an example data set
```



## Prepare Work

#### Hypotheses

> H_0: There is no difference in the DV average between IV group  

> H_a: The DV average is different for at least one IV group 



### Ensure the Data is in "long" Format 

The data must be restructured from **wide** to **long** format, so that each observation is on its own line.  All categorical variables must be declared as factors.  We also must add an distinct indicator variable.  

Name of the dataset: `data_long`


### Check the Variables


Variables:  
 
* `id_var` Name of the variable that is qunique for each unit of study (i.e. person)
* `group_IV` Name of the categorical variable (`factor`) with at least 2 levels
* `continuous_DV` Name of the continuous varaible (`dbl`) that we want to compare the averages of


## Exploratory Data Analysis

### Compute Summary Statistics



```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** Calculate summary statistics of the DV by the groups to the data to eyeball the potential effect or differences in the means.  Also check if the standard deviations and samples sizes are somewhat similar in each group.
```



```{r, eval=FALSE}
data_long %>% 
  dplyr::group_by(group_IV) %>%          # divide into groups
  furniture::table1(continuous_DV        # gives M(SD)
                    digits = 2,     
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown") 
```





### Plot the Observed Data



```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** Plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean. 
```



#### Side-by-Side Boxplots

```{r, eval=FALSE}
data_long %>% 
  ggplot(aes(x = group_IV,           # groups along the x-axis
             y = continuous_DV)) +   # DV on the y-axis
  geom_boxplot() +
  geom_point() +
  theme_bw()
```


#### Mean by Group Dot-Plot


```{r, eval=FALSE}
data_long %>% 
  ggplot(aes(x = group_IV,           # groups along the x-axis
             y = continuous_DV)) +   # DV on the y-axis 
  stat_summary() +
  theme_bw()
```


#### Violin-&-Boxplots

```{r, eval=FALSE}
data_long %>% 
  ggplot(aes(x = group_IV,           # groups along the x-axis
             y = continuous_DV)) +   # DV on the y-axis
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw()
```


## Assumption Checks


### Independent, Random Samples

The Sample(s) was drawn at random (at least as representative as possible)


```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** Nothing can be done to fix NON-representative samples and there is no statistically test to test this.
```


### Normality of the DV in each Group

> A variable is said to follow the normal distribution if it resembles the normal curve.  Specifically it is symetrical, unimodal, and bell shaped.

The continuous variable has a **NORMAL** distribution in ALL relevant populations

* Not as important if the sample is large *(Central Limit Theorem)*
* IF the sample is far from normal &/or small, might want to use a different method  



```{block type='rmdlightbulb', echo=TRUE}
**Options to judging normality:**

1. Visualization of each sample's distribution    
    + Stacked **histograms**, but is sensitive to binning choices *(number or width)*
    + Side-by-side **boxplots**, shows *median* instead of *mean* as central line
    + Seperate **QQ plots** (straight $45^\circ$ line), but is sensitive to outliers! 
    
2. Calculate **Skewness** and **Kurtosis** 
    + Divided each value by its standard error (SE)    
    + A result $\gt \pm 2$ indicates issues  
    + Best used when sample is small, *n* < 30'ish
    

3. Formal Inferential Tests for Normality  
    + Null-hypothesis: population is normally distributed
    + A $p \lt .05$ ???indicate snon-normality
    + For smaller samples, use **Shapiro-Wilk's Test** 
    + For larger samples, use **Kolmogorov-Smirnov's Test**    
```





#### Histogram by Groups    

```{r, eval=FALSE}
data_long %>% 
  ggplot(aes(continuous_var)) +           # DV along the x-axis
  geom_histogram(fill = "gray",
                 color = "black",
                 alpha = .5) +
  facet_grid(group_var ~ .,               # Panel by group
             labeller = label_both) +
  theme_bw()
```


#### QQ Plot by Group


```{r, eval=FALSE}
data_long %>% 
  ggplot(aes(sample = continuous_var)) +    # make sure to include "sample = DV"
  geom_qq() +                               
  stat_qq_line() +                          
  facet_wrap(~ group_var,                   # Panel by group
             labeller = label_both) +    
  theme_bw()
```

    

    
    
#### Skewness and Kurtosis Values    
    
```{r, eval=FALSE}
data_long %>% 
  psych::describeBy(continuous_var ~ group_var,    # Formula: DV ~ groups
                    data = .,
                    mat = TRUE,
                    digits = 2) 
```
    
    



#### Skapiro-Wilk's Test

```{r, eval=FALSE}
data_long %>% 
  dplyr::group_by(group_var) %>%                                      # By group
  dplyr::summarise(W.stat  = shapiro.test(continuous_var)$statistic,  # For the DV
                   p.value = shapiro.test(continuous_var)$p.value)    # For the DV
```
    

#### Skewness, Kurtosis, & Shapiro-Wilk's Test

```{r, eval=FALSE}
data_long %>% 
  dplyr::nest_by(group_var) %>%                                        # By group
  dplyr::summarise(summ = list(pastecs::stat.desc(data$continuous_var, # For the DV
                                                  norm = TRUE))) %>% 
  dplyr::mutate(summ = list(summ %>% 
                              unlist() %>% 
                              data.frame() %>% 
                              dplyr::rename("value" = ".") %>% 
                              tibble::rownames_to_column(var = "stat"))) %>% 
  tidyr::unnest(summ) %>% 
  dplyr::filter(stat %in% c("skewness", "skew.2SE",
                            "kurtosis", "kurt.2SE",
                            "normtest.W", "normtest.p")) %>% 
  tidyr::pivot_wider(names_from = stat,
                     values_from = value)
```

     
* `skew.2SE` 
    - Calculateed as `skew` divided by 2 times the SE of skewness
    - If > 1, then we can conclude violations to normality
    
* `kurt.2SE` 
    - Calculateed as `kurt` divided by 2 times the SE of skewness
    - If > 1, then we can conclude violations to normality 
    
    
    
    
    

### Homogeneity of Variance (HOV) in the DV between groups



All three of the Tests below have:

* $H_0$: all populations' variances are equal
* $H_1$ At least two of the populations have different variances




```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** When using **Bartlett’s test** one of the main assumptions data should be normally distributed. In the case of non-normal data, the **Levene's test** is an alternative to the Bartlett test. The **Fligner-Killeen test** is one of the many tests for homogeneity of variances which is most robust against departures from normality.
```

#### Bartlett's Test

```{r, eval=FALSE}
df_long %>% 
  bartlett.test(continuous_var ~ group_var,     # Formula: DV ~ groups
                data = .)
```

#### Leven's Test 

```{r EX_hov, eval=FALSE}
data_long %>% 
  car::leveneTest(continuous_var ~ group_var,     # Formula: DV ~ groups
                  data = .,            
                  center = "mean")     
```

#### Filgner-Killeen's Test

```{r, eval=FALSE}
df_long %>% 
  fligner.test(continuous_var ~ group_var,     # Formula: DV ~ groups
                data = .)
```






## One-way ANOVA Model



### Fit & Save the Model

The `aov_4()` function from the `afex` package fits ANOVA models *(oneway, two-way, repeated measures, and mixed design)*. It needs at least two arguments:

1. formula:  `continuous_DV ~ group_IV + (1|id_var)`  *one observation per subject and `id_var` is distinct for each subject*

2. dataset: `data = .` *we use the period to signify that the datset is being piped from above*


Here is an outline of what your syntax should look like when you **fit and save a one-way ANOVA**.  Of course you will replace the dataset name and the variable names, as well as the name you are saving it as.

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** The `aov_4()` function works on data in LONG format only.  Each observation needs to be on its one line or row with seperate variables for the group membership (categorical factor or `fct`) and the continuous measurement (numberic or `dbl`).  The `(1|id_var)` indicates there is only ONE observation PER individual.
```

```{r, eval=FALSE}
# One-way ANOVA: fit and save
aov_name <- data_long %>% 
  afex::aov_4(continuous_DV ~ group_IV + (1|id_var),       # Formula: DV ~ groups + (1| id_var)
              data = .)
```





### Basic Output - stored name of model

By running the name you saved you model under, you will get a brief set of output, including a measure of **Effect Size**.

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** The `ges` is the *generalized eta squared*.  In a one-way ANOVA, the eta-squared effect size is only one value, ie. generalized $\eta_g$ and partial $\eta_p$ are the same.
```



```{r, eval=FALSE}
# Display basic ANOVA results (includes effect size)
aov_name 
```


### Fuller Output - add $Anova on model name

To fully fill out a standard ANOVA table and compute other effect sizes, you will need a more complete set of output, including the **Sum of Squares** components, you will need to add `$Anova` at the end of the model name before running it.

```{block type='rmdlightbulb', echo=TRUE}
 **NOTE:** IGNORE the first line that starts with `(Intercept)`!  Also, the 'mean sum of squares' are not included in this table, nor is the **Total** line at the bottom of the standard ANOVA table.  You will need to manually compute these values and add them on the homework page.  Remember that `Sum of Squares (SS)` and `degrees of freedom (df)` add up, but `Mean Sum of Squreas (MS)` do not add up.  Also: `MS = SS/df` for each term.
```

```{r, eval=FALSE}
# Display fuller ANOVA results (includes sum of squares)
aov_name$Anova 
```


## Followup-tests


### Calculate Estimated Marginal Means


```{block type='rmdlightbulb', echo=TRUE}
 **NOTE:** Estimated Marginal Means (EMM) will we identical WHEN the groups are exactly the same size AND there is only ONE grouping variable.
```


```{r EX_emm, eval=FALSE}
aov_name %>% 
  emmeans::emmeans(~ group_var)        # by groups
```




### Plot Estimated Marginal Means


```{r EX_emmip, eval=FALSE}
aov_name %>% 
  emmeans::emmip(~ group_var,        # Calculate Estimated Marginal Means
                 CIs = TRUE)         # Include Confidence Intervals
```





### All Pairwise Comparisons

There are two steps to conduct all possible pairwise comparisons:

1. `emmeans::emmeans(~ group_var)` - Calculate the *Estimated Marinal Means*
2. `pairs()` - Determine if each pair is *significantly different*

Within the `pairs()` function there are several options for controling for multiple comparisons, including:

* `adjust = "none"` - Fisher's LSD
* `adjust = "tukey"` - Tukey's HSD
* `adjust = "bon"` - Bonferroni

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** Fisher's LSD approach is valid if you have 2-3 groups.  If there are 4 or more groups, Tukey's HSD adjustment is more appropriate, provided the groups are the same size.  Bonferroni's adjustment is most conservative.
```


```{r EX_tukey, eval=FALSE}
aov_name %>% 
  emmeans::emmeans(~ group_var) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "tukey")                  # Is each pair signif different?
```


### Effect Size for Pairwise Comparisons


Cohen's d like "Standardized Mean Differences"


```{r, eval=FALSE}
aov_name %>% 
  emmeans::emmeans(~ group_var) %>% 
  emmeans::eff_size(sigma = sigma(aov_name$lm),
                    edf = df.residual(aov_name$lm))
```



### Contrast Statements

There are two steps to conduct a contrast comparison:

1. `emmeans(~ group_var)` - Calculate the *Estimated Marinal Means*
2. `contrast()` - Determine if each pair is *significantly different*

Inside the contrast statement, list the named sets of linear contrast weights.  We will only be doing one-at-a-time, but we must still use a nested `list`.  

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** You must provide one weight ($c_i$) for each of the $k$ groups.  If you wish to ignore a group, that group's weight is $c_i = 0$.  The sum total of all the weights must be zero ($\sum c_i = 0$), so use positive and negative numbers.
```

```{r EX_contrast, eval=FALSE}
aov_name %>% 
  emmeans::emmeans(~ group_var) %>% 
  emmeans::contrast(list("your contrast name" = c(c_1, c_2, ... , c_k)))
```


## APA Write-up


### Methods Section

* Research Question(s)
    - Specific testable hypotheses
* Sample
    - participants
    - inclusion & exclusion criteria
* Procedure
    - setting 
    - timeline
    - intervention & placebo (if applicable)
* Variables
   - Measures used
   - DV & IV(s)
* Analytic Plan
    - Main Method Used, options
    - Assumptions & How Tested
    - Follow-up tests
    - Effect Size




### Results Section


* Descriptive Summary
    - Table 1 with sample and variables
    - Eyeball treands seen
* Main Test Results
    - Assumptions of interest
    - May need a table(s) and/or plot(s)
* Follow-up Tests
    - final conclusion

Make sure to include:
* Effect Sizes
* Direction of any Significant relationship
* Contex (not "reject the null")











