# Poison Example: 1 & 2-way ANOVA


Required Packages 

```{r}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(furniture)    # Nice tables (by our own Tyson Barrett)
library(psych)        # Descriptive helpers
library(pastecs)      # Normality Tests
library(afex)         # Analysis of Factorial Experiments
library(emmeans)      # Estimated marginal means (Least-squares means)
library(ggpubr)
```

```{r, echo=FALSE}
knitr::knit("5-functions.Rmd", output = tempfile())
```


## Prepare for Modeling

### Ensure the Data is in "long" Format 


https://www.guru99.com/r-anova-tutorial.html 

The "poison" dataset contains 48 rows and 4 variables:

* **ID**: Guinea pig identification number
* **Time**: Survival time of the animal
* **poison**: Type of poison used: factor level: 1,2 and 3
* **treat**: Type of treatment used: factor level: 1,2 and 3



Before you start to compute the ANOVA test, you need to prepare the data as follow:

1. **Import** the data from online
2. **Rename** the `id` variable
3. **Declare factors** label categories of poison type & treatment


```{r}
PATH <- "https://raw.githubusercontent.com/guru99-edu/R-Programming/master/poisons.csv"

df_survival <- read.csv(PATH) %>%
  dplyr::rename(id = X) %>% 
  dplyr::mutate(poison = factor(poison)) %>% 
  dplyr::mutate(treat = factor(treat))
```


```{r}
tibble::glimpse(df_survival)
```

```{r}
str(df_survival)
```



## Example 1) Survival Time by Poison Type

### Research Question


We want to know if there is a difference between the mean of the **survival time** according to the **type of poison** given to the Guinea pig.



#### Hypotheses

> H_0: There is no difference in survival time average between group  
> H_a: The survival time average is different for at least one group 



#### Variables

* `id` Guinea pig identification number
* `poison` poison type, 3-levels (X, independent groups, IV)
* `time` survival time, continuous (Y, dependent variable, DV)


> Note: ignore `treat` for now


### Exploratory Data Analysis


#### Summary Statistics

```{r}
df_survival %>% 
  dplyr::group_by(poison) %>%   
  furniture::table1("Survival Time" = time,
                    digits = 2,     
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown",
                    caption = "Summary of Survival Time by Type of Poisson Used")       
```

#### Plot the Observed Data



```{r, fig.cap = "Distribution of Survival Time by Poison Type: Violin and Box plots used to Compare Means and Standard Deviations"}
df_survival %>% 
  ggplot(aes(x = poison,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Poison",
       y = "Observed Survival Time") 
```


#### Summary

Things you should notice, but not a firm conclusion yet!

**SAMPLE BALANCE (*N*)**

The sample of 48 guinea pigs is balanced between the three poisons.  There are the sample number of animals in each group (*n* = 16).

**CENTER (MEANS)**  

Guinea pigs given Type 3 poison had shorter survival times (*M* = 0.28), on average, compared to Type 1 (*M* = 0.48) and Type 2 (*M* = 0.54).

**SPREAD (STANDARD DEVIATIONS)**  

There was much less variation in survival times among the 16 guinea pics given Type 3 poison (*SD* = 0.06) compared to the other types (Type 1, *SD* = 0.21; Type 2, *SD* = 0.29).







### Assumption Checks


#### Independent, Random Samples

The Sample(s) was drawn at **random** (at least as representative as possible) and the groups are **independent** of each other.

* Nothing can be done to fix NON-representative samples!
* Can not for with any statistically test
* Lack of independence requires a different model


#### Normality of the DV in each Group




```{r, fig.cap = "Distribution of Survival Time by Poison Type: Histograms used to Judge Normality"}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_histogram(binwidth = .05,
                 fill = "gray",
                 color = "black",
                 alpha = .5) +
  facet_grid(poison ~ .,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



```{r, fig.cap = "Distribution of Survival Time by Poison Type: QQ Plots used to Judge Normality"}
df_survival %>% 
  ggplot(aes(sample = time)) +    
  geom_qq() +                     
  stat_qq_line() +                
  facet_wrap(~ poison,            
             labeller = label_both) +          
  theme_bw()
```

```{r}
df_survival %>% 
  dplyr::group_by(poison) %>% 
  dplyr::summarise(skewness = psych::skew(time),
                   kurtosis = psych::kurtosi(time))
```


```{r}
df_survival %>% 
  dplyr::group_by(poison) %>% 
  dplyr::summarise(SW.stat  = shapiro.test(time)$statistic,
                   p.value = shapiro.test(time)$p.value)
```

```{r}
df_survival %>% 
  dplyr::nest_by(poison) %>% 
  dplyr::summarise(summ = list(pastecs::stat.desc(data$time, 
                                                  norm = TRUE))) %>% 
  dplyr::mutate(summ = list(summ %>% 
                              unlist() %>% 
                              data.frame() %>% 
                              dplyr::rename("value" = ".") %>% 
                              tibble::rownames_to_column(var = "stat"))) %>% 
  tidyr::unnest(summ) %>% 
  dplyr::filter(stat %in% c("skewness", "skew.2SE",
                               "kurtosis", "kurt.2SE",
                               "normtest.W", "normtest.p")) %>% 
  tidyr::pivot_wider(names_from = stat,
                     values_from = value)
```




#### Homogeneity of Variance (HOV) in the DV between groups

Since the distribution of survival time might not be normally distributed, Bartlett's test is not the best. 


```{r}
df_survival %>% 
  car::leveneTest(time ~ poison, 
                  data = .,            
                  center = "mean")     
```

```{r}
df_survival %>% 
  fligner.test(time ~ poison, 
               data = .)
```



#### Summary


**NORMALITY**

The distribution of survival time does not seem to be normal based on visual inspection.  For poison Type 2, survival time significantly deviates from normality (skew = 1.07, kurtosis = -0.08), *W* = 0.85, *p* = .015.


**HOMOGENEITY OF VARIANCE**

Survival time is significantly less variable for poison Type 3 (*SD* = 0.06), compared to the others (Type 1, *SD* = 0.21; Type 2, *SD* = 0.29), *F*(2, 45) = 8.09, *p* < .001.








### Fitting One-way ANOVA Model



```{r}
aov_poison <- df_survival %>% 
  afex::aov_4(time ~ poison + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
apa::anova_apa(aov_poison)
```



```{r}
aov_poison
```



#### Fuller Output - add $Anova on model name


```{r}
aov_poison$Anova
```

#### Summary


There is statistically significant evidence that survival time varies by poison type, *F*(2, 45) = 11.79, *p* < .001, $\eta^2$ = 0.34.


### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`

```{r}
aov_poison %>% 
  emmeans::emmeans(~ poison)         
```

```{r}
aov_poison %>% 
  emmeans::emmip(~ poison,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Poison",
       y = "Estmated Marginal Mean: Survival Time")
```






#### All Pairwise Comparisons with `pairs()`

Pairwise post hoc: Fisher's LSD adjustment for multiple comparisons

```{r}
aov_pairs <- aov_poison %>% 
  emmeans::emmeans(~ poison) %>%        
  pairs(adjust = "none")  


aov_pairs
```



#### Effect Size for Pairs



```{r}
aov_eff <- aov_poison %>% 
  emmeans::emmeans(~ poison) %>% 
  emmeans::eff_size(sigma = sigma(aov_poison$lm),
                    edf = df.residual(aov_poison$lm))

aov_eff
```




#### Contrast Statements with `emmeans::contrast()`

Compare the mean of Type 3 to the other two types.


```{r}
aov_poison %>% 
  emmeans::emmeans(~ poison) %>% 
  emmeans::contrast(list("1 & 2 vs. 3" = c(1, 1, -2)))
```




#### Summary


**PAIRS**

Survival time for Type 3 (*M* = 0.28) is is lower than type 1 (*M* = 0.62), *t*(45) = 4.61, *p* < .001, *d* = 1.63, and type 2 (*M* = 0.54), *t*(45) = 3.62, *p* < .001, *d* = 1.28.  There is no evidence between types 1 & 2, *t*(45) = 0.99, *p* = .328, *d* = 0.35.


**CONTRAST**

Survival time for Type 3 (*M* = 0.28) is is lower than  both type 1 (*M* = 0.62) and type 2 (*M* = 0.54), *t*(45) = 4.75, p < .001.



### APA Write Up

> Note: There are many unkowns regarding the situation of this experiment and can not be include in this write-up.

#### Methods Section


To investigate potency of poisons, 48 guinea pigs were randomized to three types of poison (*n* = 16) and time till mortality was measured.  Survival time (dependent variable) was then subjected to a 1-way independent groups analysis of variance (ANOVA) comparing the three types of poison (independent variable).  Prior to analysis, the assumptions of normality and homogeneity of variance were assessed with with Shapiro-Wilks and Levene's Test respectively. The significant omnibus effect was investigated further with pairwise and appropriate contrast comparisons to determine which of the types different on survival time.  Fisher's method was used for all pairwise tests and effects sizes are given as Cohen's d-like standardized mean differences.



#### Results Section

The distribution of survival time did not seem to be normal based on visual inspection.  For poison Type 2, survival time significantly deviates from normality (skew = 1.07, kurtosis = -0.08), *W* = 0.85, *p* = .015. Survival time also was significantly less variable for poison Type 3 (*SD* = 0.06), compared to the others (Type 1, *SD* = 0.21; Type 2, *SD* = 0.29), *F*(2, 45) = 8.09, *p* < .001.  The ANOVA was still conducted and found statistically significant evidence that survival time do indeed varry by poison type, *F*(2, 45) = 11.79, *p* < .001, $\eta^2$ = 0.34.  Specifically, survival time for Type 3 (*M* = 0.28) was shorter than type 1 (*M* = 0.62), *t*(45) = 4.61, *p* < .001, *d* = 1.63, and type 2 (*M* = 0.54), *t*(45) = 3.62, *p* < .001, *d* = 1.28.  There was no evidence of a difference in survival time between types 1 & 2, *t*(45) = 0.99, *p* = .328, *d* = 0.35.


> Here is a publication worthy plot

```{r}
aov_poison_label <- dplyr::left_join(data.frame(aov_pairs), 
                 data.frame(aov_eff), 
                 by = "contrast") %>% 
  tidyr::separate(col = contrast,
                  into = c("start", "stop")) %>% 
  dplyr::mutate_at(vars(start, stop),
                   stringr::str_sub,
                   start = -1) %>% 
  dplyr::mutate(lab = glue::glue("d = {round(effect.size, 2)}, {pval_label_apa(p.value)}"))

aov_poison %>% 
  emmeans::emmeans(~ poison) %>% 
  data.frame() %>% 
  ggplot(aes(x = poison,
             y = emmean)) +
  geom_errorbar(aes(ymin = emmean - SE,
                    ymax = emmean + SE),
                width = .3) +
  geom_point() +
  geom_line(aes(group = 1)) +
  theme_bw() +
  labs(x = "Poison Type",
       y = "Estimated MArginal Mean Survival Time") +
  ggpubr::geom_bracket(data = aov_poison_label %>% dplyr::filter(p.value < .06),
               aes(xmin = start,
                   xmax = stop,
                   label = lab),
               y.position = 0.7,
               step.increase = -0.15,
               type = "text") 
```





## Example 2) Survival Time by Treatment Type

### Parepare for Analysis

Our objective is to test the following assumption:

> H_0: There is no difference in survival time average between group  

> H_a: The survival time average is different for at least one group.


In other words, you want to know if there is a statistical difference between the mean of the survival time according to the type of treatment given to the Guinea pig.

**Variables:**

* `id` Guinea pig identification number
* `treat` treatment type, 4-level IV (independent groups)
* `time` survival time, continuous DV (Y, outcome)



> Note: ignore `poison` for now



#### Compute Summary Statistics

Second, check the summary statistics for each of the $k=4$ groups.

```{r}
df_survival %>% 
  dplyr::group_by(treat) %>%   # divide into groups
  furniture::table1("Survival Time" = time,
                    digits = 2,      # gives M(SD)
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown",
                    caption = "Summary of Survival Time by Type of Treatment Used")       
```


**Interpretion:**




#### Plot the Raw Data

Plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.  I have added a red diamond on the mean survival time for each type of poison.

```{r}
df_survival %>% 
  ggplot(aes(x = treat,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Observed Survival Time")
```


#### Assumption Check: Normality




```{r}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_histogram(binwidth = .10,
                 color = "black",
                 fill = "gray",
                 alpha = .5) +
  facet_grid(treat ~ .,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



#### Assumption Check: HOV

Levene's Test of HOV

```{r}
df_survival %>% 
  car::leveneTest(time ~ treat, 
                  data = .,            
                  center = "mean")     
```





### Fitting One-way ANOVA Model



```{r}
aov_treat <- df_survival %>% 
  afex::aov_4(time ~ treat + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
aov_treat
```



#### Fuller Output - add $Anova on model name


```{r}
aov_treat$Anova
```




### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`

```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat)         # Calculate Estimated Marginal Means
```

```{r}
aov_treat %>% 
  emmeans::emmip(~ treat,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Estmated Marginal Mean: Survival Time")
```



#### All Pairwise Comparisons with `pairs()`

Pairwise post hoc: Fisher's LSD adjustment for multiple comparisons

```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "none")                  # Is each pair signif different?
```


#### Contrast Statements with `emmeans::contrast()`


```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat) %>% 
  emmeans::contrast(list("B vs. Rest" = c(1, -3, 1, 1)))
```




## Example 3) Survival Time by Both Poisson & Treatment Type


### Parepare for Analysis

Our objective is to test the following assumption:

> H_0: There is no difference in survival time average between group  
> H_a: The survival time average is different for at least one group


In other words, you want to know if there is a statistical difference between the mean of the survival time according to BOTH  type of poisons and treatment given to the Guinea pig.



**Variables:**

* `id` Guinea pig identification number
* `poison` poison type, 3-level IV (independent groups)
* `treat` treatment type, 4-level IV (independent groups)
* `time` survival time, continuous DV (Y, outcome)





#### Compute Summary Statistics

Second, check the summary statistics for each of the $k$ groups.

```{r}
df_survival %>% 
  dplyr::group_by(treat, poison) %>%   # divide into groups
  dplyr::summarise(n = n(),
                   mean = mean(time),
                   sd = sd(time)) %>% 
  gt::gt(caption = "Summary of Survival Time by Type of Treatment Used")
```




#### Plot the Raw Data

Third, plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.  I have added a red diamond on the mean survival time for each type of poison.

```{r}
df_survival %>% 
  ggplot(aes(x = treat,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Observed Survival Time") +
  facet_wrap(~ poison, labeller = label_both)
```


#### Assumption Check: Normality


```{r}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_density(fill = "gray",
               alpha = .5) +
  facet_grid(treat ~ poison,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



#### Assumption Check: HOV

Levene's Test of HOV

```{r}
df_survival %>% 
  car::leveneTest(time ~ treat*poison, 
                  data = .,            
                  center = "mean")     
```







### Fitting Two-way ANOVA Model



```{r}
aov_survival2 <- df_survival %>% 
  afex::aov_4(time ~ poison*treat + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
aov_survival2
```



#### Fuller Output - add $Anova on model name


```{r}
aov_survival2$Anova
```




### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`





#### All Pairwise Comparisons for Main Effects of Poison

```{r}
aov_survival2 %>% 
  emmeans::emmeans(~ poison) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "none")                  # Is each pair signif different?
```



```{r}
aov_survival2 %>% 
  emmeans::emmip( ~ poison,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Poison",
       y = "Estmated Marginal Mean: Survival Time")
```



#### All Pairwise Comparisons for Main Effects of Treatment

```{r}
aov_survival2 %>% 
  emmeans::emmeans(~ treat) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "tukey")                  # Is each pair signif different?
```




```{r}
aov_survival2 %>% 
  emmeans::emmip( ~ treat,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of 
       Treatment",
       y = "Estmated Marginal Mean: Survival Time")
```







