# Poison Example: 1 & 2-way ANOVA


Required Packages 

```{r}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(furniture)    # Nice tables (by our own Tyson Barrett)
library(afex)         # Analysis of Factorial Experiments
library(emmeans)      # Estimated marginal means (Least-squares means)
library(readxl)       # Necessary for reading in an example data set
```

## Prepare for Modeling

### Ensure the Data is in "long" Format 


https://www.guru99.com/r-anova-tutorial.html 

The "poison" dataset contains 48 rows and 4 variables:

* ID: Guinea pig identification number
* Time: Survival time of the animal
* poison: Type of poison used: factor level: 1,2 and 3
* treat: Type of treatment used: factor level: 1,2 and 3



Before you start to compute the ANOVA test, you need to prepare the data as follow:

1. Import the data from online
2. Rename the "id" variable
3. Declare factors: poison type & treatment


```{r}
PATH <- "https://raw.githubusercontent.com/guru99-edu/R-Programming/master/poisons.csv"

df_survival <- read.csv(PATH) %>%
  dplyr::rename(id = X) %>% 
  dplyr::mutate(poison = factor(poison)) %>% 
  dplyr::mutate(treat = factor(treat))
```


```{r}
tibble::glimpse(df_survival)
```

```{r}
str(df_survival)
```



## Example 1) Survival Time by Poison Type

### Parepare for Analysis

Our objective is to test the following assumption:

H_0: There is no difference in survival time average between group
H_a: The survival time average is different for at least one group.


In other words, you want to know if there is a statistical difference between the mean of the survival time according to the type of poison given to the Guinea pig.

Variables:

* `id` (`id_var`) Name of the variable that is qunique for each unit of study (i.e. person)
* `poison` (`group_IV`) Name of the categorical variable (`factor`) with at least 2 levels
* `time` (`continuous_DV`) Name of the continuous varaible (`dbl`) that we want to compare the averages of

> Note: ignore `treat` for now



#### Compute Summary Statistics

Second, check the summary statistics for each of the $k$ groups.

```{r}
# Raw data: summary table
df_survival %>% 
  dplyr::group_by(poison) %>%   # divide into groups
  furniture::table1("Survival Time" = time,
                    digits = 2,      # gives M(SD)
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown",
                    caption = "Summary of Survival Time by Type of Poisson Used")       
```





#### Plot the Raw Data

Third, plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.  I have added a red diamond on the mean survival time for each type of poison.

```{r}
df_survival %>% 
  ggplot(aes(x = poison,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Observed Survival Time") 
```






#### Assumption Check: Normality


```{r}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_density(fill = "gray",
               alpha = .5) +
  facet_grid(poison ~ .,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



#### Assumption Check: HOV

Levene's Test of HOV

```{r}
df_survival %>% 
  car::leveneTest(time ~ poison, 
                  data = .,            
                  center = "mean")     
```






### Fitting One-way ANOVA Model



```{r}
aov_poison <- df_survival %>% 
  afex::aov_4(time ~ poison + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
aov_poison
```



#### Fuller Output - add $Anova on model name


```{r}
aov_poison$Anova
```




### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`

```{r}
aov_poison %>% 
  emmeans::emmeans(~ poison)         # Calculate Estimated Marginal Means
```

```{r}
aov_poison %>% 
  emmeans::emmip(~ poison,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Poison",
       y = "Estmated Marginal Mean: Survival Time")
```



#### All Pairwise Comparisons with `pairs()`

Pairwise post hoc: Fisher's LSD adjustment for multiple comparisons

```{r}
aov_poison %>% 
  emmeans::emmeans(~ poison) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "none")                  # Is each pair signif different?
```


#### Contrast Statements with `emmeans::contrast()`


```{r}
aov_poison %>% 
  emmeans::emmeans(~ poison) %>% 
  emmeans::contrast(list("1 & 2 vs. 3" = c(1, 1, -2)))
```










## Example 2) Survival Time by Treatment Type

### Parepare for Analysis

Our objective is to test the following assumption:

H_0: There is no difference in survival time average between group
H_a: The survival time average is different for at least one group.


In other words, you want to know if there is a statistical difference between the mean of the survival time according to the type of treatment given to the Guinea pig.

Variables:

* `id` (`id_var`) Name of the variable that is qunique for each unit of study (i.e. person)
* `treat` (`group_IV`) Name of the categorical variable (`factor`) with at least 2 levels
* `time` (`continuous_DV`) Name of the continuous varaible (`dbl`) that we want to compare the averages of

> Note: ignore `poison` for now



#### Compute Summary Statistics

Second, check the summary statistics for each of the $k$ groups.

```{r}
# Raw data: summary table
df_survival %>% 
  dplyr::group_by(treat) %>%   # divide into groups
  furniture::table1("Survival Time" = time,
                    digits = 2,      # gives M(SD)
                    na.rm = FALSE,
                    total = TRUE,
                    output = "markdown",
                    caption = "Summary of Survival Time by Type of Treatment Used")       
```




#### Plot the Raw Data

Third, plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.  I have added a red diamond on the mean survival time for each type of poison.

```{r}
df_survival %>% 
  ggplot(aes(x = treat,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Observed Survival Time")
```


#### Assumption Check: Normality


```{r}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_density(fill = "gray",
               alpha = .5) +
  facet_grid(treat ~ .,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



#### Assumption Check: HOV

Levene's Test of HOV

```{r}
df_survival %>% 
  car::leveneTest(time ~ treat, 
                  data = .,            
                  center = "mean")     
```





### Fitting One-way ANOVA Model



```{r}
aov_treat <- df_survival %>% 
  afex::aov_4(time ~ treat + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
aov_treat
```



#### Fuller Output - add $Anova on model name


```{r}
aov_treat$Anova
```




### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`

```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat)         # Calculate Estimated Marginal Means
```

```{r}
aov_treat %>% 
  emmeans::emmip(~ treat,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Estmated Marginal Mean: Survival Time")
```



#### All Pairwise Comparisons with `pairs()`

Pairwise post hoc: Fisher's LSD adjustment for multiple comparisons

```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "none")                  # Is each pair signif different?
```


#### Contrast Statements with `emmeans::contrast()`


```{r}
aov_treat %>% 
  emmeans::emmeans(~ treat) %>% 
  emmeans::contrast(list("B vs. Rest" = c(1, -3, 1, 1)))
```




## Example 3) Survival Time by Both Poisson & Treatment Type


### Parepare for Analysis

Our objective is to test the following assumption:

H_0: There is no difference in survival time average between group
H_a: The survival time average is different for at least one group.


In other words, you want to know if there is a statistical difference between the mean of the survival time according to the type of treatment given to the Guinea pig.

Variables:

* `id` (`id_var`) Name of the variable that is qunique for each unit of study (i.e. person)
* `treat` & `poison` (`group_IV`) Name of the categorical variable (`factor`) with at least 2 levels
* `time` (`continuous_DV`) Name of the continuous varaible (`dbl`) that we want to compare the averages of

> Note: ignore `poison` for now



#### Compute Summary Statistics

Second, check the summary statistics for each of the $k$ groups.

```{r}
df_survival %>% 
  dplyr::group_by(treat, poison) %>%   # divide into groups
  dplyr::summarise(n = n(),
                   mean = mean(time),
                   sd = sd(time)) %>% 
  gt::gt(caption = "Summary of Survival Time by Type of Treatment Used")
```




#### Plot the Raw Data

Third, plot the data to eyeball the potential effect.  Remember the center line in each box represents the median, not the mean.  I have added a red diamond on the mean survival time for each type of poison.

```{r}
df_survival %>% 
  ggplot(aes(x = treat,
             y = time)) + 
  geom_violin(fill = "gray") +
  geom_boxplot(width = .07,
               fill = "white") +
  geom_jitter(position = position_jitter(0.21)) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               color = "red",
               size = 5) +
  theme_bw() +
  labs(x = "Type of Treatment",
       y = "Observed Survival Time") +
  facet_wrap(~ poison, labeller = label_both)
```


#### Assumption Check: Normality


```{r}
df_survival %>% 
  ggplot(aes(time)) + 
  geom_density(fill = "gray",
               alpha = .5) +
  facet_grid(treat ~ poison,
             labeller = label_both) +
  theme_bw() +
  labs(x = "Observed Survival Time")
```



#### Assumption Check: HOV

Levene's Test of HOV

```{r}
df_survival %>% 
  car::leveneTest(time ~ treat*poison, 
                  data = .,            
                  center = "mean")     
```







### Fitting Two-way ANOVA Model



```{r}
aov_survival2 <- df_survival %>% 
  afex::aov_4(time ~ poison*treat + (1|id),
              data = .)
```



#### Basic Output - stored name of model

```{r}
aov_survival2
```



#### Fuller Output - add $Anova on model name


```{r}
aov_survival2$Anova
```




### Followup-tests

#### Estimated Marginal Means with `emmeans::emmeans()`





#### All Pairwise Comparisons for Main Effects of Poison

```{r}
aov_survival2 %>% 
  emmeans::emmeans(~ poison) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "none")                  # Is each pair signif different?
```



```{r}
aov_survival2 %>% 
  emmeans::emmip( ~ poison,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of Poison",
       y = "Estmated Marginal Mean: Survival Time")
```



#### All Pairwise Comparisons for Main Effects of Treatment

```{r}
aov_survival2 %>% 
  emmeans::emmeans(~ treat) %>%        # Calculate Estimated Marginal Means
  pairs(adjust = "tukey")                  # Is each pair signif different?
```




```{r}
aov_survival2 %>% 
  emmeans::emmip( ~ treat,
                 CIs = TRUE) +
  theme_bw() +
  labs(x = "Type of 
       Treatment",
       y = "Estmated Marginal Mean: Survival Time")
```







