#Mixed Design ANOVA

Required Packages 

```{r load_libraries}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(furniture)    # Nice tables (by our own Tyson Barrett)
library(afex)         # needed for ANOVA, emmeans is loaded automatically.
library(multcomp)     # for advanced control for multiple testing/Type 1 error
```

## Tutorial - Fitting Mixed Design ANOVA Models with `afex::aov_4()`

The `aov_4()` function from the `afex` package fits ANOVA models (oneway, two-way, repeated measures, and mixed design). It needs at least two arguments:

1. formula:  `continuous_var ~ group_var + (RM_var|id_var)`  *one observation per subject for each level of the `RMvar`, so each `id_var` has multiple lines for each subject, each subject can only belong to exactly one group./*

2. dataset: `data = .` *we use the period to signify that the datset is being piped from above*


Here is an outline of what your syntax should look like when you **fit and save a Mixed ANOVA**.  Of course you will replace the dataset name and the variable names, as well as the name you are saving it as.

```{block type='rmdlightbulb', echo=TRUE}
**NOTE:** The `aov_4()` function works on data in LONG format only.  Each observation needs to be on its one line or row with seperate variables for the group membership (categorical factor or `fct`) and the continuous measurement (numberic or `dbl`).
```


```{r EX_Mixaov_4, eval=FALSE}
# RM ANOVA: fit and save
aov_name <- data_name %>% 
  afex::aov_4(continuous_var ~ group_var + (RM_var|id_var),
              data = .)
```

## Words Recalled Data Example (Chapter 16, section A)


### Data Prep

I input the data as a `tribble` which saves it as a `data.frame` and then cleaned up a few of the important variables.
```{r, warning=FALSE, message=FALSE}
d <- tibble::tribble(
  ~ID, ~depression, ~word_type, ~words_recalled,
    1,           0,          1,              20,
    2,           0,          1,              16,
    3,           0,          1,               8,
    4,           1,          1,              17,
    5,           1,          1,              15,
    6,           1,          1,              10,
    1,           0,          2,              21,
    2,           0,          2,              18,
    3,           0,          2,               7,
    4,           1,          2,              15,
    5,           1,          2,              10,
    6,           1,          2,               4,
    1,           0,          3,              17,
    2,           0,          3,              11,
    3,           0,          3,               4,
    4,           1,          3,              18,
    5,           1,          3,              13,
    6,           1,          3,              10) %>%
  mutate(depression = factor(depression, 
                             labels = c("Not Depressed",
                                        "Depressed"))) %>%
  mutate(word_type = factor(word_type,
                            labels = c("Neutral", "Positive", "Negative")))
d
```

### One-Way RM ANOVA

First, let's ignore the depression variable and just look at the repeated measures. We do this by using `afex::aov_4()` and then the `summary()` functions as shown below.

```{r}
oneway <- d %>%
  afex::aov_4(words_recalled ~ 1 + (word_type|ID),
              data = .)
summary(oneway)
```

Here, we see a number of pieces of information, including the sums of squares, F-statistic, and p-value. The p-value suggests that there is not an effect of `word_type` here (p = .342).

Let's plot this using a spaghetti plot.

```{r, warning=FALSE, message=FALSE}
d %>%
  ggplot(aes(word_type, words_recalled, group = ID)) +
    geom_line() +
    geom_point()
```

But we wonder if depression has an effect on the number of words recalled, and it may interact with word_type. Let's see what that looks like.

```{r, warning=FALSE, message=FALSE}
d %>%
  ggplot(aes(word_type, words_recalled, group = ID, 
             color = depression, 
             shape = depression, 
             linetype = depression)) +
    geom_line() +
    geom_point() +
    facet_wrap(~depression)
```

Definitely looks like the effect of word_type depends on whether the individual has depression or not. To add a between subjects factor to a repeated measures ANOVA, we are now doing mixed ANOVA (both between and within subjects factors). 

### Mixed ANOVA

To run a mixed ANOVA, use `afex::aov_4()` and then the `summary()` functions again but this type with the between subjects factor included.

```{r}
mixed_anova <- d %>%
  afex::aov_4(words_recalled ~ depression + (word_type|ID),
              data = .)
summary(mixed_anova)
```

The output provides us with a clear significant interaction shown in the first table. Our previous plot helps illustrate what this interaction is telling us about the patterns. However, it is often useful to show what the groups are doing, not the individuals (especially if your sample size is larger than 20). To do that, we can use:

```{r}
mixed_anova %>%
  emmeans::emmip(depression ~ word_type)
```

From this, we can tell that there is very little difference with neutral words, but large differences for positive and negative words. Specifically, depressed individuals struggle much more at recalling positive words than non-depressed individuals and depressed individuals do better at recalling negative words than non-depressed individuals.

## Conclusion

Overall, this is a good approach to using mixed ANOVA:

1. Look at the repeated measures first, and then,
2. Include any potentially important between-subjects factors.